Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-03-17 demystify.tex
updates

Most PhD programs will require at least \textbf{two LORs}. Having a strong letter from an internationally recognized researcher will \emph{greatly strengthen} your application. In fact, a strong letter from a well-known researcher can outweigh the lack of publications (or low GPA or unknown school).

However, obtaining letters from famous people
can be challenging for international students, who might not have much interactions with such experts. So it is fine to have a letter from people (e.g., profs, researchers, postdocs who mentored you) that know you well enough to talk about \emph{your research experience and capabilities}. Many students get letters from supervisors from companies where they did internships or are working at. It is OK as long as it is a research-based personalized letter (once again, we are talking about PhD applications, not MS). Again, the emphasis here is \emph{research}, i.e., the letters should describe your research experiences and potentials. Letters focusing your course work or non-research projects at a company won't carry much weight.

@@

\paragraph{Open House} Most schools will have \textbf{Open House} or visit events, which are a great resource to learn about the school, department, faculty, research, living, etc. Even if you can't come in person, you should attend virtually and meet with individual faculty.
During the Open House, you get a chance to talk to individual faculty and current students.  Take notes of faculty who make you excited, count those that are taking in new students (if they meet you, likely they are considering new students!).  Talk to students about their advisers, the dept, the area, funding situation etc.  Ask about anything you want to determine that they deserve \emph{you}.

@@

Also, do not forget to update and thank your LoR writters and others who have supported you throughout your admission journey.

\paragraph{If you do not get admitted} If you do not get admitted to any schools or don't want to go to the ones that admit you, try again next time.  Graduate admission can involve randomness and noise.  Don't bother asking for feedback, you will not likely get any.  In the meantime, you can work on improving your profile, e.g., get more research experiences, publish more papers, improve your connections for better LoR writers, etc.

You can also consider applying to MS programs, which are typically easier to get in; but you need to pay.  If you get in an MS program at a school of your choice, you can contact professors to work with them. If you do well, you can ask the professor to support you to convert to PhD.  This is not uncommon for many students.

@@

\section{Why did I get rejected?}\label{sec:why-rejected}
Many possible reasons!  Other than not having a strong profile (e.g., research potentials, GPA, LoRs, SOP), here are some that you might not expect:

\begin{itemize}

\item \emph{You are not a good fit to the program}.  You could have an excellent profile, but if you are interested in a research area that the program does not have, you might not be admitted.
Similarly, if there is no faculty who is willing to advise you (e.g., they are on sabbatical, personal leave, or already have too many students already), you will not be admitted.  This is actually good for you, as you don't want to be in a program where no one can advise you.

@@

\item  \emph{Your English scores (e.g., IELTS or TOEFL) are not high enough}.  While professors might not care much about these, there is a certain minimum set by the college or the university that you need to pass to get admitted, especially for TA funding (e.g., you need to be able to communicate well with students as a TA).  Thus, while professors are willing to argue for your case, they might be reluctant to go against the standard requirement of the university on English scores.  Note this is mostly for international students, as domestic ones (or those from these \href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-GMU#standard-tests-waiver-eligible-countries}{countries}) do not have to take these tests.

\end{itemize}

@@

\section{Buying Computer Equipment}\label{sec:buying-equipment} understandably students get very excited about their upcoming PhD journal and want to buy new computer equipment and electronics to prepare. However, you should check with your adviser first.  Many profs. have funding to buy you computers and other equipments (e.g., software, books, keyboards, headphones, tablets, etc). Many CS programs also provide some budget or computer equipments to incoming PhD students (e.g., a laptop).

@@

@@

\paragraph{Transcripts} You do not need to get your transcripts evaluated (which can be a hassle and inaccurate).  You can just send your official transcripts directly to the school you're applying to.

@@

\paragraph{Standing out \autoref{sec:improve-your-chance}} There are \emph{very few} domestic applications compared to international ones, i.e., domestic students are minority (\autoref{sec:urm}) in the CS PhD application pool. Many US universities want to increase the number of domestic students in their programs (and as mentioned above, there are specific fellowships for domestic students).
That makes you stand out and can help your case.

>>>>

Most PhD programs will require at least \textbf{two LORs}. A good LoR letter should have detail, enthusiasm, and specific examples of the applicant's achievements, all of which paint a compelling picture of the candidate to the adcom.

Having a strong letter from an internationally recognized researcher will \emph{greatly strengthen} your application. In fact, a strong letter from a well-known researcher can outweigh the lack of publications (or low GPA or unknown school).
However, obtaining letters from famous people
can be challenging for international students, who might not have much interactions with such experts. So it is fine to have a letter from people (e.g., profs, researchers, postdocs who mentored you) that know you well enough to talk about \emph{your research experience and capabilities}. Many students get letters from supervisors from companies where they did internships or are working at. It is OK as long as it is a research-based personalized letter (once again, we are talking about PhD applications, not MS). Again, the emphasis here is \emph{research}, i.e., the letters should describe your research experiences and potentials. Letters focusing your course work or non-research projects at a company won't carry much weight.

@@

\paragraph{Open House} Most schools have \emph{Open House} or \emph{Visit Day} events, which are a great resource to learn about the school, department, faculty, research, living, etc.
\textbf{Advice}: Even if you can't come in person, you should attend virtually and meet with individual faculty. During the event, you get a chance to learn more about the program, talk to individual faculty and current students.  Take notes of faculty who make you excited, count those that are taking in new students (if they meet you, likely they are considering new students!).  Talk to students about their advisers, the dept, the area, funding situation etc.  Ask about anything you want to determine that they deserve \emph{you}.

@@

Also, do not forget to update and thank your LoR writters and others who have supported you through this process.

\paragraph{If you do not get admitted} If you do not get admitted to any schools or don't want to go to the ones that admit you, try again next time.  Graduate admission can involve randomness and noise. In the meantime, you can work on improving your profile, e.g., get more research experiences, publish more papers, improve your connections for better LoR writers, etc. See rejection reasons (\autoref{sec:why-rejected}) and additional advice ()\autoref{sec:improve-your-chance}).

You can also consider applying to MS programs, which are typically easier to get in; but you need to pay.  If you get in an MS program at a school of your choice, you can contact professors to work with them. If you do well, you can ask the professor to support you to convert to PhD.

@@

\section{Why did I get rejected?}\label{sec:why-rejected}
Many possible reasons!  Other than not having a strong profile (e.g., research potentials, GPA, LoRs, SOP, poor interviews), here are some that you might not expect:

\paragraph{Not a good fit}  You could have an excellent profile, but if you are interested in a research area that the program does not have, you might not be admitted.
Similarly, if there is no faculty who is willing to advise you (e.g., they are on sabbatical, personal leave, do not have sufficient fundings, or already have too many students already), you will not be admitted.  This is actually good for you, as you don't want to be in a program where no one can advise you.

@@

\paragraph{Low English exam scores (e.g., IELTS or TOEFL)}  While professors might not care much about these, there might be a certain minimum set by the college or the university that you need to pass to get admitted, especially for TA funding (e.g., you need to communicate well with students as a TA).  Thus, while professors are willing to argue for your case, they might be reluctant to go against the standard requirement of the university on English scores.  Note this is for international students, as domestic ones (or those from these \href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-GMU#standard-tests-waiver-eligible-countries}{countries}) do not have to take these tests.

\paragraph{Overqualified or Lack Interest}  This might be paradoxical, however if the adcom believes that you are likely to get admitted and go to a better program (i.e., they are just your ``safety'' schools), they might not be so enthusiastic to admit you and want to save the spot for someone else. Note though it does happen, for most good schools this is rare and you should not assume this is the case.

A related reason is that you did not show enough interest in the program.
For example, your SOP did not mention any faculty or research area at the program, or you did not respond to professors interested in interviewing you (which might be considered unprofessional and burn bridges).

\paragraph{Red flags in Your Background} Plagiarism, cheating, or other academic dishonesty are red flags. Another one is if you have a history of jumping from one program to another (e.g., you have been in multiple different programs in short time). If you think you have red flags, you should address them in your SOP.
In general, these things are rare, but they do happen and cause concern to the adcom.

  %\item \emph{You did not make a good impression}.  For example, you might have contacted a professor and they did not like your email.  Or you might have met a professor (e.g., at a conference) and they did not like your presentation.  This is rare, but it happens.

@@

\section{Buying Computer Equipment}\label{sec:buying-equipment}

Understandably students get excited about their upcoming PhD journey and want to buy new computer equipment and electronics to prepare. However, you should check with your adviser first.  Many profs. have funding to buy you computers and other equipments (e.g., software, books, keyboards, headphones, tablets, etc). Many CS programs also provide some budget or computer equipments to incoming PhD students (e.g., a laptop).

@@

\paragraph{Standing out \autoref{sec:improve-your-chance}} There are \emph{few} domestic applications compared to international ones, i.e., domestic students are minority (\autoref{sec:urm}) in the CS PhD application pool. Many US universities want to increase the number of domestic students in their programs (and as mentioned above, there are specific fellowships for domestic students).
That makes you stand out and can help your case.

@@

\paragraph{Transcripts} You do not need to get your transcripts evaluated/translated (which can be a hassle and inaccurate).  You can just send your official transcripts directly to the school you're applying to.

@@
